---
title: 'Unit 2 Case Study: Identifying At-Risk Students'
author: "Dr. Shaun Kellogg"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_depth: '5'
    toc_float: yes
    code_folding: show
    code_download: TRUE
name: ''
editor_options:
  markdown:
    wrap: 72
bibliography: lit/references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. PREPARE

In Unit 2, we pick up where we left off and learn to apply some basic
machine learning techniques to help us understand how a predictive model
might be developed and tested as part of an early warning system to
identify students at risk of failing. Specifically, we will make a very
crude first attempt at developing a testing a logistic regression and a
random forest model that can (we hope!) accurately predict whether a
student will pass or fail and online course. Unlike Unit 1, where we
were interested in identifying factors from data collected throughout
the course that might help explain why students earned the grade they
did, we are instead interested identifying students who may be at risk
before it is too late to intervene.

### 1a. Load Libraries

#### tidymodels 📦

[![](img/tidymodels.svg){width="20%"}](https://www.tidymodels.org)

The [tidymodels](https://www.tidymodels.org) package is a "meta-package"
for modeling and statistical analysis that shares the underlying design
philosophy, grammar, and data structures of the
[tidyverse](https://www.tidyverse.org/). It includes a core set of
packages that are loaded on startup and contains tools for:

-   data splitting and pre-processing;
-   model selection, tuning, and evaluation;
-   feature selection and variable importance estimation;
-   as well as other functionality.

#### [**Your Turn**]{style="color: green;"} **⤵**

In addition to the {tidymodels} package, we'll also be using the
{tidyverse} packages we learned about in Unit 1, and the {ranger}
package we'll be using for our random forest model in Part 4.

Use the code chunk below to load these three packages:

```{r}
library(tidymodels)
library(tidyverse)
library(ranger)
```

### b. Import & Inspect Data

For this case study, we will again be working again with data from the
online science courses introduced in Unit 1. This data is located in the
data folder and named `processed-data.csv`. Fortunately, our data has
already been largely wrangled. This will save us quite a bit of time,
which we'll need to help wrap our heads around some supervised machine
learning basics.

Run the code chunk below to read this data into your R environment as an
object named `processed_data` and let's take a quick look at data we may
use in our predictive model:

```{r}
processed_data <- read_csv("data/processed-data.csv")

processed_data
```

#### [**Your Turn**]{style="color: green;"} **⤵**

In addition to students' gender and their pre-course survey responses
assessing three aspects of student motivation, we will also incorporate
gradebook data collected during the course that we will also use to help
"train" a predictive model for identifying students at risk of failing.

Use the code chunk below to read in the .csv file named "grade-book.csv"
located in the data folder, assign to a new object named `grade_book`,
and use a method of your choosing to answer the questions that follow:

```{r}
#Your Code Here
grade_book <- read_csv("data/grade-book.csv")

grade_book
```

1.  How many observations and variables are in our `grade_book` dataset?
    And roughly how many observations are there per student?

    -   YOUR RESPONSE HERE

2.  How might this data be used in our model to help predict student at
    risk of failing the course?

    -   YOUR RESPONSE HERE

3.  What else do you notice about this data set?

    -   YOUR RESPONSE HERE

## 2. WRANGLE

In general, data wrangling involves some combination of cleaning,
reshaping, transforming, and merging data (Wickham & Grolemund, 2017).
The importance of data wrangling is difficult to overstate, as it
involves the initial steps of going from raw data to a dataset that can
be explored and modeled (Krumm et al, 2018). In Part 2, we focus on the
the following wrangling processes to:

a.  **Create an Outcome Variable**. We create a binary variable outcome
    we are interested in predicting, i.e. pass/fail.

b.  **Convert Character Stings to Factors.** We convert variables stored
    as character strings to factors expected by our models.

c.  **Select Predictors.** We narrow down our data set to specific
    predictors of interest.

### a. Create Outcome Variable

Since we are interested in developing a predictive model that can
predict whether a student is at risk of failing a course, and so we can
intervene before that happens, we need an outcome variable that let's us
know if they have either passed or failed.

Let's combine the hopefully familiar `mutate()` function with the
`if_else()` function also from the {dplyr} package to create a new
variable called `at_risk` which indicates "no" for students whose
`final_grade` is greater than or equal to 66.7% (NC State's cutoff for a
D-) and "yes" if it is not.

```{r}
course_data <- processed_data %>% 
  mutate(at_risk = if_else(final_grade >= 66.7, "no", "yes")) 

course_data
```

### b. Convert to Factors

While inspecting the data, you may have noticed that one of our
predictors of interest and our `at_risk` outcome variable are stored as
character \<chr\> data types. For creating models, it is better to have
qualitative variables like `gender` and `at_risk` encoded as factors
instead of character strings. Factors store data as categorical
variables, each with its own levels. Because categorical variables are
used in statistical models differently than continuous variables,
storing these data as factors ensures that the modeling functions will
treat them correctly.

To do so, we once again use the `mutate()` function but instead of
creating a new variable, we will use the `as_factor()` function to
convert `gender` and `at_risk` from a character to a factor and save it
as a variable of the same name, effectively replacing the old variables
with new ones:

```{r}
course_data <- course_data |>
  mutate(gender = as_factor(gender), 
         at_risk = as_factor(at_risk))

course_data
```

**Note:** Be sure to pay attention to the fact that in addition to
assigning our modified `gender` variable to a variable of the same name
using the `=` operator, we also assigned our modified `course_data` to
an object of the same name using the `<-` operator. That is, we
effectively overwrite the old `course_data` that included `gender`
stored as a character data type to a new `course_data` object with
`gender` stored as a factor.

### c. Select Predictors

As you've probably noticed, there is a lot of great information in our
`course_data` - but we won't, and shouldn't, include all of it in our
predictive model. Indeed, Estrellado et al. @estrellado2020data point
out that for statistical reasons and as a good general practice, it's
better to select a few variables of interest because:

> At a certain point, adding more variables will *appear* to make your
> analysis more accurate, but will in fact obscure the truth from you.

#### [**Your Turn**]{style="color: green;"} **⤵**

Complete the code chunk below to "select" (hint, hint) `at_risk` for our
outcome variable, as well as the following variables we will use in our
predictive model:

-   student_id
-   course_id
-   at_risk
-   gender
-   course_interest
-   perceived_competence
-   utility_value

```{r}
course_data <- course_data |>
  select(student_id, 
         course_id, 
         at_risk, 
         gender, 
         course_interest, 
         perceived_competence,
         utility_value)

course_data
```

## 3. Explore

As noted by Krumm et al. (2018), exploratory data analysis often
involves some combination of data visualization and feature engineering.
In Part 3, we will create a quick visualization to help us spot any
issues with our data and engineer new features that we will use in our
predictive models. Specifically, in Part 3 we will:

a.  **Examine Outcomes** by taking a quick look at `at_risk` outcome
    variable of interest as well as our other predictors to spot any
    issues that may hang up out models.

b.  **Engineer Predictors** by creating some new variables we think will
    be predictive of students at risk, such as performance on
    assignments on during the first half of the course.

### a. Count proportions

Before we move on to feature engineering, let's take a quick look at the
proportion of students in our final data set that were identified as
at-risk. To do so, run the following code chunk to take a count of the
number of students labeled "yes" or "no" of being `at_risk`, and then
create a new variable called proportion that takes the number `n` of
each and divides by the total number:

```{r}
course_data |>
  count(at_risk) |>
  mutate(proportion = n/sum(n))
```

As you can see, roughly 18% of students in our data set were identified
as "yes" for being at risk. And since these are final grades we are
working with, these student were not only at risk of failing, but did
indeed fail the course. You may also notice we have a number of students
with missing grades, which is something we will have to address prior to
analysis.

#### [**Your Turn**]{style="color: green;"} **⤵**

Alternatively, we could have created a quick visualization to help us
get a better sense of the outcomes we are trying to predict and to spot
any potential issues we might run into during analysis.

Complete the following code chunk to create a simple bar plot
illustrating the number of students who passed, failed, or have missing
data:

```{r}
course_data |>
  ggplot(aes(x = at_risk)) +
  geom_bar()
```

#### Remove Cases

As you may have noticed from the outputs above, some students do not
have a final grade and some are missing survey data as well. Since the
models we will be using are not designed to deal with missing data, we
will need to remove cases with missing data.

Let's use the `drop_na()` function also from the {dplyr} package to
remove observations with missing data and reassign to our `course_data`
object:

```{r}
course_data <- course_data |>
  drop_na()

course_data
```

#### [**Your Turn**]{style="color: green;"} **⤵**

Finally, you have probably noticed that we wrote a lot more code than
necessary in order to illustrate different wrangling processes to get
our data ready for analysis.

To reduce all the redundancy caused by demonstrating each step
separately above, complete the following code below by using the new
`|>` or old `%>%` [pipe
operators](https://medium.com/analytics-vidhya/how-to-use-the-new-pipe-in-r-4-1-f7bf748a465d)
to chain the `mutate()`, `select()`, and `drop_na()` functions together
to wrangle our data for analysis:

```{r}
course_data <- processed_data |>
  mutate(at_risk = if_else(final_grade >= 66.7, "no", "yes"), # 
         at_risk = as_factor(at_risk), #
         gender = as_factor(gender)) |> # 
  select(student_id, # 
         course_id, 
         at_risk, 
         gender, 
         course_interest, 
         perceived_competence,
         utility_value) |>
  drop_na() # 

course_data
```

Now inspect your data again and make sure it looks like expected:

```{r}
course_data
```

**Hint:** You should see a total of 503 observations and 7 variables
including 1 outcome variable named `at_risk`, 1 student identifier named
`student_id`.

### b. Feature Engineering

As defined by Krumm, Means, and Bienkowski (2018) in *Learning Analytics
Goes to School*:

> **Feature engineering** is the process of creating new variables
> within a dataset, which goes above and beyond the work of recoding and
> rescaling variables.

The authors note that feature engineering draws on substantive knowledge
from theory or practice, experience with a particular data system, and
general experience in data-intensive research. Moreover, these features
can be used not only in machine learning models, but also in
visualizations and tables comprising descriptive statistics.

Though not often discussed, feature engineering is an important element
of data-intensive research that can generate new insights and improve
predictive models. Indeed, [an earlier
attempt](https://datascienceineducation.com/c14.html) using this data
without feature engineering predicted students' passing (or not passing)
the course with only around 75% accuracy. One goal of this case study is
to improve upon these results creating by some new variable using data
that we think may improve our model.

#### Student Performance Features

In addition to information about student gender and motivation collected
prior to the course, we will also incorporate into our model student
performance data on course assignments. And since we are interested in
developing a predictive model that that can be used to intervene before
students actually fail the course, we will develop new features based on
student performance on the first 20 assignments completed during the
first half of the course. Specifically, we'll create the following three
"features" from the gradebook data we imported earlier:

-   The overall percent of points earned (*across* first 20 assignments)
-   The variability (in standard deviation units) in the percent earned
    (*between* assignments)
-   The number of assignments for which students earned 100% of the
    possible points

#### Group & Slice Data

The {dplyr} package has a handy `slice()` function that that allows us
to select, remove, and duplicate rows in a dataset based the rows in
which we specify.

For example, run the following code chunk to select rows 1 through 20
from our `grade_book` data.

```{r}
slice(grade_book, 1:20)
```

While this helps us retrieve the first 20 assignments for the first
student, it doesn't help us with the hundreds of other students in our
data set!

Fortunately, the {dplyr} package also has the extremely useful
`group_by()` function which allows us to perform other {tidyverse}
functions like `slice()`, `mutate()`, and `summarize()` by groups that
we specify as arguments.

For example, run the following code to *group* our data by `student_id`
and `course_id` using the `group_by()` function, and then we'll use the
`slice()` function again to select the first 20 assignment, but this
time for the first 20 assigments each course the student completed,
instead of just the first 20 assignments for the first student:

```{r}
grade_book |> 
  group_by(student_id, course_id) |>
  slice(1:20)
```

You'll notice the data have significantly different dimensions now.
We'll have to take some steps to further process our `grade_book` data.
In doing so, we'll engineer some features.

#### [Your Turn]{style="color: green;"} ⤵

Now let's create a new `grade_book` data frame using what we just
learned to:

-   group our observations by `student_id` and `course_id`;
-   slice our data frame to include only the first 20 assigments for
    each student in each course;
-   create a new variable called `percent_earned` that equals the
    `points_earned` divided by `points_possible`; and,
-   remove rows with any missing data

Complete the following code chunk:

```{r}

grade_book <- grade_book |> 
  group_by(student_id, course_id) |>
  slice(1:20) |>
  mutate(percent_earned = points_earned/points_possible) |>
  drop_na()

grade_book
```

**Note:** If you completed this correctly, you should have a data frame
with 10,522 observations and 6 variables.

#### Create New Variables

Finally, let's create three features from the gradebook data:

-   The overall percent of points earned
-   The variability (in standard deviation units) in the percent earned
    (*between* assignments)
-   The number of assignments for which students earned 100% of the
    possible points

You can probably imagine others; you're welcome to explore adding those,
too.

We'll use the `summarize()` function instead of the `mutate()` function
to do this, and then we'll select just the variables needed to join our
data and :

```{r}
grade_book <- grade_book |>
  summarize(overall_percent = sum(points_earned) / sum(points_possible),
            variability = sd(percent_earned),
            n_100 = sum(percent_earned == 1)) |>
  select(student_id, 
         course_id, 
         overall_percent, 
         variability, 
         n_100)

grade_book
```

We have one last step before we can get to modeling (`grade_book`) -
joining this data with all of the other data (`course_data`).

```{r}
course_data <- inner_join(course_data, grade_book)
```

#### [Your Turn]{style="color: green;"} ⤵

Let's talk a look at the joined data to make sure everything is looking
as we intend it to. Inspect the data using the code chunk below and
answer the following questions:

```{r}
course_data
```

1.  What does our data frame now have 503 observations and 10 variables?
    -   YOUR RESPONSE HERE
2.  Why did we use the `summarize()` function above instead of the
    `mutate()` function to create our new features?
    -   YOUR RESPONSE HERE

## 4. MODEL

Recall from our readings that there are two general types of modeling
approaches: unsupervised and supervised machine learning. In Part 4, we
focused on supervised learning models, which are used to quantify
relationships between features (e.g., motivation and performance) and a
known outcome (e.g., passing or failing a course). These models can be
used for statistical inference, as illustrated in Unit 1, or prediction
a we'll illustrate in this section. Specifically, in Part 4 we will
learn how to:

a.  **Split Data** into a training and test set that will be used to
    develop a predictive model.

b.  **Create a "Recipe"** for our predictive model and learn how to deal
    with nominal data that we would like to use as predictors.

c.  **Fit Models** to our training set using logistic regression and
    random forest models.

d.  **Check Model Accuracy** on our test set to see how well our model
    can "predict" our outcome of interest.

### a. Split Data

The authors of Data Science in Education Using R @estrellado2020data
remind us that:

> At its core, machine learning is the process of "showing" your
> statistical model only some of the data at once and training the model
> to predict accurately on that training dataset (this is the "learning"
> part of machine learning). Then, the model as developed on the
> training data is shown new data - data you had all along, but hid from
> your computer initially - and you see how well the model that you
> developed on the training data performs on this new testing data.
> Eventually, you might use the model on entirely new data.

It is therefore common when beginning a modeling project to [separate
the data set](https://bookdown.org/max/FES/data-splitting.html) into two
partitions:

-   The *training set* is used to estimate develop and compare models,
    feature engineering techniques, tune models, etc.

-   The *test set* is held in reserve until the end of the project, at
    which point there should only be one or two models under serious
    consideration. It is used as an unbiased source for measuring final
    model performance.

There are different ways to create these partitions of the data and
there is no uniform guideline for determining how much data should be
set aside for testing. The proportion of data can be driven by many
factors, including the size of the original pool of samples and the
total number of predictors. 

After you decide how much to set aside, the most common approach for
actually partitioning your data is to use a random sample. For our
purposes, we'll use random sampling to select 25% for the test set and
use the remainder for the training set, which are the defaults for the
{[rsample](https://tidymodels.github.io/rsample/)} package.

Additionally, since random sampling uses random numbers, it is important
to set the random number seed. This ensures that the random numbers can
be reproduced at a later time (if needed).

The function `initial_split()` function from the {rsample} package takes
the original data and saves the information on how to make the
partitions.

Run the following code to set the random number seed and make our
initial data split:

```{r}
set.seed(586)

course_split <- initial_split(course_data, 
                              strata = at_risk)
```

Note that we used the `strata =` argument, which conducts a stratified
split. This ensures that, despite the imbalance we noticed in our
`at_risk` variable, our training and test data sets will keep roughly
the same proportions of at-risk students as in the original data. 

#### [**Your Turn**]{style="color: green;"} **⤵**

Type `course_split` into the code chunk below, run, and answer the
question that follows?

```{r}
course_split
```

1.  How many observations should we expect to see in our training and
    test sets respectively?

    -   YOUR RESPONSE HERE

#### Create a training and test set

The {rsample} package has two aptly named functions for creating a
training and testing data set called `training()` and `testing()`
respectively.

Run the following code to split the data into our training and test data
sets:

```{r}
train_data <- training(course_split)

test_data  <- testing(course_split)
```

#### [**Your Turn**]{style="color: green;"} **⤵**

Now take a look at the training and testing sets we just created:

```{r}
train_data

test_data
```

Next, recycle the code from above to check to see that the proportion of
at-risk students in our training and test data are close to those in our
overall `course_data`:

```{r}
train_data |>
  count(at_risk) |>
  mutate(proportion = n/sum(n))

test_data |>
  count(at_risk) |>
  mutate(proportion = n/sum(n))
```

Now answer the following questions:

1.  Do the number of observations in each set match your expectations?
    Why?

    -   YOUR RESPONSE HERE

2.  Do the proportion of at-risk students in each set match your
    expectations? Why?

    -   YOUR RESPONSE HERE

### b. **Create a Recipe**

In this section, we introduce another tidymodels package,
[recipes](https://recipes.tidymodels.org/), which is designed to help
you prepare your data *before* training your model. Recipes are built as
a series of preprocessing steps, such as:

-   converting qualitative predictors to indicator variables (also known
    as dummy variables),

-   transforming data to be on a different scale (e.g., taking the
    logarithm of a variable),

-   transforming whole groups of predictors together,

-   extracting key features from raw variables (e.g., getting the day of
    the week out of a date variable),

and so on. If you are familiar with R's formula interface, a lot of this
might sound familiar and like what a formula already does. Recipes can
be used to do many of the same things, but they have a much wider range
of possibilities.

#### Add a formula

To get started, let's create a recipe for a simple logistic regression
model. Before training the model, we can use a recipe to add a few
predictors and conduct some preprocessing required by the model.

The [`recipe()` function](https://recipes.tidymodels.org/reference/recipe.html) as
we used it here has two arguments:

-   A **formula**. Any variable on the left-hand side of the tilde (`~`)
    is considered the model outcome (`at_risk` in our case). On the
    right-hand side of the tilde are the predictors. Variables may be
    listed by name, or you can use the dot (`.`) to indicate all other
    variables as predictors.

-   The **data**. A recipe is associated with the data set used to
    create the model. This will typically be the *training* set,
    so `data = train_data` here. Naming a data set doesn't actually
    change the data itself; it is only used to catalog the names of the
    variables and their types, like factors, integers, dates, etc.

Let's create our very first recipe using `at_risk` as our outcome
variable; `course_interest` and `gender` and `overall_percent` as
predictors; and `train_data` as our data to train:

```{r}
lr_recipe_1 <- recipe(at_risk ~ course_interest + gender + overall_percent,
                      data = train_data)
```

Now let's take a quick peek at our recipe and create a quick summary of
our recipe using the `summary()` function:

```{r}
lr_recipe_1

summary(lr_recipe_1)
```

You can see that our recipe has four ingredients including three
predictors and 1 outcome, just as expected.

#### Create Dummy Variables

Because we'll be using a simple logistic regression model, variables
like `gender` will need to be coded as [dummy
variables](https://bookdown.org/max/FES/creating-dummy-variables-for-unordered-categories.html).
Dummy coding means transforming a variable with multiple categories into
new variables, where each binary variable indicates the presence and
absence of each category. For example, gender will be recoded to
gender_f, where 1 indicates female and 0 indicates male.

Unlike the standard model formula methods in R, a recipe **does not**
automatically create these dummy variables for you; you'll need to tell
your recipe to add this step. This is for two reasons. First, many
models do not require [numeric
predictors](https://bookdown.org/max/FES/categorical-trees.html), so
dummy variables may not always be preferred. Second, recipes can also be
used for purposes outside of modeling, where non-dummy versions of the
variables may work better. For example, you may want to make a table or
a plot with a variable as a single factor.

For these reasons, we need to explicitly tell recipes to create dummy
variables using `step_dummy()`. Let's add this to our recipe and include
`all_nominal_predictors()` to tell our recipe to change all of our
factor variables to dummy variables:

```{r}
lr_recipe_1 <- 
  recipe(at_risk ~ course_interest + gender + overall_percent,
         data = train_data) |>
  step_dummy(all_nominal_predictors())

lr_recipe_1

summary(lr_recipe_1)
```

#### Create a kitchen sink recipe

Before training our model, let's create a second recipe just for
contrast that includes all of our predictors.

Run the following code to add all our predictors to our new recipe:

```{r}
lr_recipe_2 <- 
  recipe(at_risk ~ course_interest + gender + overall_percent + 
           perceived_competence + utility_value + variability + n_100, 
         data = train_data) |>
  step_dummy(all_nominal_predictors())

lr_recipe_2

summary(lr_recipe_2)
```

### c. Fit a Logistic Regression Model

With tidymodels, we start building a model by specifying the *functional
form* of the model that we want using the [[parsnip]
package](https://tidymodels.github.io/parsnip/). Since our outcome is
binary, the model type we will use is "[logistic
regression](https://parsnip.tidymodels.org/reference/logistic_reg.html)".
We can declare this with `logistic_reg()` and assign to an object we
will later use in our workflow:

```{r}
lr_mod <- logistic_reg()
```

That is pretty underwhelming since, on its own, it doesn't really do
much. However, now that the type of model has been specified, a method
for *fitting* or training the model can be stated using the **engine**.

#### Start your engine

The engine value is often a mash-up of different packages that can be
used to fit or train the model as well as the estimation method. For
example, we will use "glm" a generalized linear model for binary
outcomes and default for logistic regression in the {parsnip} package.

Run the following code to finish specifying our model:

```{r}
lr_mod <- 
  logistic_reg() %>% 
  set_engine("glm")
```

#### Add to workflow

We will want to use our recipes created earlier across several steps as
we train and test our model. To simplify this process, we can use
a *model workflow*, which pairs a model and recipe together.

This is a straightforward approach because different recipes are often
needed for different models, so when a model and recipe are bundled, it
becomes easier to train and test *workflows*.

We'll use the {[workflows](https://workflows.tidymodels.org/)} package
from tidymodels to bundle our parsnip model (`lr_mod`) with our first
recipe (`lr_recipe_1`).

```{r}
lr_workflow <- 
  workflow() %>% 
  add_model(lr_mod) %>% 
  add_recipe(lr_recipe_1)
```

#### Fit model to training set

Now that we have a single workflow that can be used to prepare the
recipe and train the model from the resulting predictors, we can use the
`fit()` function to fit our model to our `train_data`. And again, we set
a random number seed to ensure that if we run this same code again, we
will get the same results in terms of the data partition:

```{r}
set.seed(586)

lr_fit <- 
  lr_workflow %>% 
  fit(data = train_data)
```

This `lr_fit` object has the finalized recipe and fitted model objects
inside. To extract the model fit from the workflow, we will use the
helper functions `extract_fit_parsnip()`. Here we pull the fitted model
object then use the `broom::tidy()` function to get a tidy tibble of
model coefficients:

```{r}
lr_fit %>% 
  extract_fit_parsnip() %>% 
  tidy()
```

Among the predictors included in our model, there are no statistically
significant (i.e., p-value \< 0.05) relationships between our predictors
and being identified as at-risk according to our definition, which
doesn't bode too well for our model but let's proceed with testing
anyways.

#### Test the model

Now that we've fit our model to our training data, we're FINALLY ready
to test our model on the data we set aside in the beginning. Just to
recap the steps that led to this moment, however, recall that we:

1.  identified the model to use (`lr_mod`);

2.  created a preprocessing recipe consisting or predictors and outcome
    (`lr_recipe_1`);

3.  bundled the model and recipe into a workflow (`lr_workflow`); and

4.  trained our workflow using a single call to `fit()`.

The next step is to use the trained workflow (`lr_fit`) to predict
outcomes for our unseen test data, which we will do with the
function `predict()`. The `predict()` method applies the recipe to the
new data, then passes them to the fitted model.

```{r}

predict(lr_fit, test_data)
```

Because our outcome variable `at_risk` here is a factor, the output
from `predict()` returns the predicted class: `no` versus `yes`. Not a
super useful output to be honest though.

Fortunately there is an `augment()` function we can use with our
`lr_fir` model and `test_data` to save them together:

Let's use this function, save as `lr_predictions`:

```{r}
lr_predictions <- augment(lr_fit, test_data)
```

#### [**Your Turn**]{style="color: green;"} **⤵**

Take a quick look at `lr_predictions` in the code chunk below and answer
the question that follows?

```{r}
lr_predictions
```

Was our model successful at predicting any students who were at-risk?
How do you know?

-   YOUR RESPONSE HERE

**Hint:** Scroll to the end of the data frame and take a look at our
original `at_risk` outcome and the `.pred_class` variable which shows
the predicted outcomes.

### d. Check model accuracy

As you probably noticed, just looking at the `lr_predictions` object is
not the easiest way to check for model accuracy. Fortunately, the
{[yardstick}](https://tidymodels.github.io/yardstick/) package has an
`accuracy()` function for looking at the overall classification
accuracy, which uses the hard class predictions to measure performance.

Hard class predictions tell us whether our model
predicted `yes` or `no` for each student in the `.pred_class` column, as
well as the estimating a probability. A simple 50% probability cutoff is
used to categorize a student as at risk. For example, student `47448`
had a `.pred_no` probability of 0.8854332 and `.pred_yes` probability of
0.11456677 and so was classified as `no` for not being for `at_risk`.

Run the following code to `select()` these variables followed by the
`accuracy()` function to see how frequently our prediction matched our
observed data:

```{r}
lr_predictions |>
  select(at_risk, .pred_class) |>
  accuracy(truth = at_risk, .pred_class)
```

Overall it looks like our model was correct 81% of the time, which is
not too bad but we'll see in a second is not good enough to serve it's
intended purpose of identifying at-risk students.

Another way to check model accuracy is with the `conf_mat()` function
from {yardstick} for creating a confusion matrix. Recall from our course
text Learning Analytics Goes to School that a confusion matrix is simply
a 2 × 2 table that lists the number of true-negatives, false-negatives,
true-positives, and false-positives.

Run the follow code to create a confusion matrix for our logistic
regression predictions:

```{r}
lr_predictions %>%
conf_mat(at_risk, .pred_class)
```

As you can see our model accurately predicted all students as "no" for
at risk 103 times, but inaccurately predicted 24 students a "no" for at
risk when they were actually "yes" in our `test_data`. Overall our model
was 81% (103/127) accurate, but it achieved this by simply labeling
everyone as "no" for at risk. And since roughly 82% of our students were
actually "no," this is clearly not a great prediction model.

#### [**Your Turn**]{style="color: green;"} **⤵**

Recycle our code from above to create a workflow for `lr_recipe_2` and
test our "kitchen sink" model on our test data. **Hint:** You can
accomplish this by simply copying and pasting and changing a single
character from above.

```{r, warning=FALSE}

# create workflow
lr_workflow <- 
  workflow() %>% 
  add_model(lr_mod) %>% 
  add_recipe(lr_recipe_2)


# set seeed
set.seed(586)

# fit model to workflow
lr_fit <- 
  lr_workflow %>% 
  fit(data = train_data)

# extract model estimates
lr_fit %>% 
  extract_fit_parsnip() %>% 
  tidy()

# get predictions 
lr_predictions <- augment(lr_fit, test_data)

# check overall accuracy 
lr_predictions %>%
  select(at_risk, .pred_class) %>%
  accuracy(truth = at_risk, .pred_class)

# create a confusion matrix
lr_predictions %>%
conf_mat(at_risk, .pred_class)
```

Does this model perform any better? How do you know?

-   YOUR RESPONSE HERE

### d. Fit a Random Forest Model

[Random forest models](https://en.wikipedia.org/wiki/Random_forest)
are [ensembles](https://en.wikipedia.org/wiki/Ensemble_learning) of
[decision trees](https://en.wikipedia.org/wiki/Decision_tree). Each of
those terms is linked because there is a dense amount of information
required to understand what each of those terms really means. For now,
however, them main thing to know is that:

> One of the benefits of a random forest model is that it is very low
> maintenance; it requires very little preprocessing of the data and the
> default parameters tend to give reasonable results.

We'll also be using the {ranger} pacakage which provides a fast
implementation of Random Forest models and is particularly suited for
high dimensional data, e.g., models with a lot of predictors and
features. This package supports supervised learning approaches including
classification, regression, survival and probability prediction.

Let's create a recipe for our `course_data` data that includes all our
predictors.

```{r}
library(ranger)

rf_recipe <- 
  recipe(at_risk ~ gender + course_interest + perceived_competence + 
           utility_value + variability + n_100 + overall_percent, 
         data = train_data)

rf_recipe

summary(rf_recipe)
```

To fit a random forest model on the training set, we'll use
the {[parsnip](https://tidymodels.github.io/parsnip/)} package again
along with the
[ranger](https://cran.r-project.org/web/packages/ranger/index.html)
engine. We'll also include the `set_mode()` function to specify our
model as "classification" rather than "regression."

Recall from our Learning Analytics Goes to School that supervised
machined learning, or predictive modeling, involves two broad
approaches: classification and regression. Classification algorithms
model categorical outcomes (e.g., yes or no outcomes like with our at
risk data). Regression algorithms characterize continuous outcomes
(e.g., test scores).

Run the following code to create our random forest model for our
training data:

```{r}
rf_mod <- 
  rand_forest(trees = 5000) %>% 
  set_engine("ranger", importance = "impurity") %>% 
  set_mode("classification")
```

Now let's combine our model and recipe to create our new random forest
workflow:

```{r}
rf_workflow <- 
  workflow() %>% 
  add_model(rf_mod) %>% 
  add_recipe(rf_recipe)
```

And fit our model and recipe to our training data:

```{r}
set.seed(586)

rf_fit <- 
  rf_workflow %>% 
  fit(data = train_data)
```

Gather our random forest predictions:

```{r}

rf_predictions <- augment(rf_fit, test_data)

rf_predictions
```

And check our model's overall accuracy:

```{r}
rf_predictions %>%
  select(at_risk, .pred_class) %>%
  accuracy(truth = at_risk, .pred_class)
```

And create a confusion matrix to see where it did and did not make
accurate predictions:

```{r}
rf_predictions |>
conf_mat(at_risk, .pred_class)
```

Unfortunately, our random forest model wasn't much better at predicting
students who were defined as "at risk."

Even more unfortunate, if this had been a real-world situation, only 1
of the 24 students who failed the course would have received additional
support. Clearly we'd need to build a better model. In this case, more
data would definitely help, but it's likely we're missing some important
information about students that is predictive of their success.

#### [**Your Turn**]{style="color: green;"} **⤵**

Try creating your own recipe and logistic regression or random forest
model and see how it performs against the three that we just created.
Feel free to experiment with variables we excluded, like specific survey
items. There is also a file in your data folder
called `sci-mo-with-text.csv` that includes data about students
participation in discussion forums, including the number of posts and
psychometric properties about their language used. You can read more
about that data in [Chapter 14 of
DSIEUR](https://datascienceineducation.com/c14.html#data-sources-3).

I recommend creating and using an R script to develop your model. Use
the code chunk below to record your final model:

```{r}

```

Now answer the following questions?

1.  How did your model do compared to others?

    -   YOUR RESPONSE HERE

2.  What information might be useful to know about students prior to the
    course start that might be useful for improving our mode?

    -   YOUR RESPONSE HERE

## 5. COMMUNICATE

In this case study, we focused applying some basic machine learning
techniques to help us understand how a predictive model used in early
warning systems might actually be developed and tested. Specifically, we
made a very crude first attempt at developing a model using machine
learning techniques that were not terribly great at accurately predict
whether a student is likely to pass or fail and online course.

Below, add a few notes in response to the following prompts:

1.  One thing I took away from this learning lab that I found especially
    useful:

    -   YOUR RESPONSE HERE

2.  One thing I want to learn more about:

    -   YOUR RESPONSE HERE

To "turn in" your work, you can click the "Knit" icon at the top of the
file, or click the dropdown arrow next to it and select "Knit top HTML".
This will create a report in your Files pane that serves as a record of
your completed assignment and that be opened in a browser or shared on
the web.

### Congratulations!