---
title: 'Laser Lab 4 - pdf Final Workflow'
author: "Jeanne McClure"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## 0. Introduction

Provide a brief overview or case study.

-   Include your R1: *research questions!*

## 1. Prepare

Load Packages

```{r warning =FALSE, message=FALSE}

#Load necessary packages
library(tidyverse)
library(here)
```

```{r eval = FALSE}

# install Latex - this may take a few minutes. After it is installed you do not need to keep it on your file. Doing this in a separate script would be better.

install.packages("tinytex")
tinytex::install_tinytex()
```

## 2. Wrangle

### a. *Import Data*

#### Data Source #1: Log Data

Log-trace data is data generated from our interactions with digital
technologies, such as archived data from social media postings. In
education, an increasingly common source of log-trace data is that
generated from interactions with LMS and other digital tools.

The data we will use has already been "wrangled" quite a bit and is a
summary type of log-trace data: the number of minutes students spent on
the course. While this data type is fairly straightforward, there are
even more complex sources of log-trace data out there (e.g., time stamps
associated with when students started and stopped accessing the course).

Let's use the `read_csv()` function from {readr} to import our
`log-data.csv` file directly from our data folder and name this data set
`time_spent`, to help us to quickly recollect what function it serves in
this analysis:

```{r}
#load with read_csv package
time_spent <- read_csv("~/RProj22/foundation_labs_2022/foundation_lab_2/data/log-data.csv")

#read in data_to_explore
data_to_explore <- read_csv(here("data", "data_to_explore.csv"))
```



## 3. Explore



#### A. TABLE SUMMARY
```{r}
#install package if this is first time using skimr


#load library
library(skimr)

#skim data
skim(data_to_explore)

```

## B. TIDY to EXPLORE
```{r}
# using the `select()` and `filter()` functions. In the code chunk below,look at descriptive for just `proportion_earned` , `time_spent` and `gender`, but only for the "OcnA" and "PhysA" subjects.

data_to_explore %>% 
  select(c('subject', 'gender', 'proportion_earned', 'time_spent')) %>% 
  filter(subject == "OcnA" | subject == "PhysA") %>%
  skim()
```


#### B. DATA VIZ

ggplot grammar - with layers


##### layers - Scatter Plot

Basic graph 
1. data
2. aes
3. geom

```{r}
#layer 1: add data and aesthetics mapping 
ggplot(data_to_explore, #<<
       aes(x = time_spent_hours, 
           y = proportion_earned)) +
#layer 2: +  geom function type
  geom_point() #<<
```

Add **Scale ** with different color for enrollment status.

```{r}
#layer 1: add data and aesthetics mapping 
#layer 3: add color scale by type
ggplot(data_to_explore, 
       aes(x = time_spent_hours, 
           y = proportion_earned,
           color = enrollment_status)) + #<<
#layer 2: +  geom function type
  geom_point()
```


Add another layer with **labs* labeling the title
```{r}
#layer 1: add data and aesthetics mapping 
#layer 3: add color scale by type
ggplot(data_to_explore, 
       aes(x = time_spent_hours, 
           y = proportion_earned,
           color = enrollment_status)) +
#layer 2: +  geom function type
  geom_point() +
#layer 4: add lables
  labs(title="How Time Spent on Course LMS is Related to Points Earned in the Course", x="Time Spent (Hours)", y = "Proportion of Points Earned")
```


Add the **facet** layer
```{r echo = TRUE, eval=FALSE}
#layer 1: add data and aesthetics mapping 
#layer 3: add color scale by type
ggplot(data_to_explore, aes(x = time_spent_hours, y = proportion_earned, color = enrollment_status)) +
#layer 2: +  geom function type
  geom_point() +
#layer 4: add lables
    x_lab(title="How Time Spent on Course LMS is Related to Points Earned in the Course", 
       x="Time Spent (Hours)",
       y = "Proportion of Points Earned")
#layer 5: add facet wrap
  facet_wrap(~ subject) 
```




##### layers- Histogram

Create a basic histogram using the 'geom_hist()' function
```{r hist1, message=FALSE, warning=FALSE}

# Layer 1: add data and aesthetic mapping
data_to_explore %>% #<<
  ggplot(aes(x = time_spent_hours)) +
# layer 2: add histogram geom
  geom_histogram()
```


Change bin size

```{r hist2, message=FALSE, warning=FALSE}
# Layer 1: add data and aesthetic mapping
data_to_explore %>% 
  ggplot(aes(x = time_spent_hours)) +
# layer 2: add histogram geom 
# layer 3a: add bin size
  geom_histogram(bins = 10)
```


Add color label to make bins stand out

```{r hist3,message=FALSE, warning=FALSE}
# Layer 1: add data and aesthetic mapping
data_to_explore %>% 
  ggplot(aes(x = time_spent_hours)) +
# layer 2: add histogram geom 
# layer 3a: add bin size
#layer 3b: add color
  geom_histogram(bins = 10,
                 fill = "red", 
                 colour = "black") 
```


Add labels and add a theme for a clean aesthetic


```{r hist4, message=FALSE, warning=FALSE}
# Layer 1: add data and aesthetic mapping
data_to_explore %>% 
  ggplot(aes(x = time_spent_hours)) +
# layer 2: add histogram geom 
# layer 3a: add bin size
# layer 3b: add color
  geom_histogram(bins = 10, fill = "red", colour = "black")+
#layer 4: add Labels
  labs(title="Time Spent on LMS histogram plot",x="Time Spent(hours)", y = "Count")+
  theme_classic()
```

Create a histogram with density plot

```{r hist5, message=FALSE, warning=FALSE}
data_to_explore%>%
  ggplot(aes(x=time_spent_hours)) +
  geom_histogram(aes(y=..density..), colour="black", fill="white", bins = 10)+
 geom_density(alpha=.2, fill="#FF6666") 
  labs(title="Time Spent on LMS histogram/density plot",x="Time Spent(hours)", y = "Density")+
  theme_classic()
```



## 4. Model
Quantify the insights using mathematical models.

#### A. MATHMATICAL

Does time spent predict grade earned?

```{r}
# Use linear regression model
lm(proportion_earned ~ time_spent_hours, 
   data = data_to_explore)
```

```{r}
# Add predictor variable for science
lm(proportion_earned ~ time_spent_hours + int, 
   data = data_to_explore)

# save the model
m1 <- lm(proportion_earned ~ time_spent_hours + int, data = data_to_explore)

```

Run a summary model for the model you just created called, `m1.`
```{r}
#run the summary
summary(m1)

```

```{r}
#install apaTables if this is your first time - do you remember how?

#load packages
library(apaTables)
# use the {apaTables} package to create a nice regression table that could be used for later publication.
apa.reg.table(m1, filename = "lm-table.doc")

```

## Communicate

**RQ1**: Do males outperform females in online STEM courses?

```{r warning=FALSE, message=FALSE}
data_viz <- data_to_explore %>%
  select(subject, gender, proportion_earned) %>%  # reduced 
  mutate(subject = recode(subject, 
                          "AnPhA" = "Anatomy",
                          "BioA" = "Biology", 
                          "FrScA" = "Forensics", 
                          "OcnA" =  "Oceanography", 
                          "PhysA" = "Physics")) %>%
  mutate(grade = proportion_earned * 100) %>%
  # filter(!is.na(gender)) %>%
  na.omit() %>% # removed all NAs instead of just those for gender
  group_by(subject, gender) %>% # grouped by subject and gender
  summarise(grade = mean(grade), 
            sd = sd(grade))# calculated mean and sd for grade and saved as grade again  

  ggplot(data_viz, aes(x = subject, y = grade, 
                          fill = gender)) +
  geom_bar(stat = "identity", 
           position = position_dodge()) +
  labs(title = "Do Males out-preform Females in online STEM courses?",
       caption = "Online STEM course performance, why is there still a gender gap?",
       y = "Average Grade",
       x = "Online STEM Course")
```

1. **Purpose** -  

2. **Methods** - 

3. **Findings** - 

4. **Discussion** - 



