---
title: "Foundations lab 4 - Data Products"
author: "Jeanne McClure"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Prepare

Install and load library's needed for analysis.
```{r warning=FALSE, message=FALSE}
#load libraries
library(tidyverse)
library(here)
```

## 2. Wrangle

Read in previously saved data
```{r messages=FALSE, warning=FALSE}
# read in data to explore 
data_to_explore <- read_csv(here("data", "data_to_explore.csv"))

```


## Explore
Looked at insights using ggplot
What insights did we find?

## Model
Quantify the insights using mathematical models. As highlighted in.[Chapter 3 of Data Science in Education Using
R](https://datascienceineducation.com/c03.html), the.**Model**.step of the data science process entails "using statistical models, from simple to complex, to understand trends and patterns in the data." The authors note that while descriptive statistics and data visualization during the**Explore**step can help us to identify patterns and relationships in our data, statistical models can be used to help us determine if relationships, patterns and trends are actually meaningful. 

##### A. Correlation Matrix

As highlighted in @macfadyen2010, scatter plots are a useful initial approach for identifying potential correlational trends between variables under investigation, but to further interrogate the significance of selected variables as indicators of student achievement, a simple correlation analysis of each variable with student final grade can be conducted.

There are two efficient ways to create correlation matrices, one that is best for internal use, and one that is best for inclusion in a manuscript. The {corrr} package provides a way to create a correlation matrix in a {tidyverse}-friendly way. Like for the {skimr} package, it can take as little as a line of code to create a correlation matrix. If not familiar, a correlation matrix is a table that presents how *all of the variables* are related to *all of the other variables*.



```{r messages=FALSE, warning=FALSE}
#install corrr package if this is your first time

# read in library
library(corrr)
```

Look and see if there is a simple correlation between time spent and student achievement.
```{r messages = FALSE, warning=FALSE}
data_to_explore %>% 
  select(proportion_earned, time_spent_hours) %>%
  correlate()
```

For printing purposes,the `fashion()` function can be added for converting a correlation data frame into a matrix with the correlations cleanly formatted (leading zeros removed; spaced for signs) and the diagonal (or any NA) left blank.

```{r messages=FALSE, warning=FALSE}
#add fashion function
data_to_explore %>% 
  select(proportion_earned, time_spent_hours) %>% 
  correlate() %>% 
  rearrange() %>%
  shave() %>%
  fashion()
```

If time allows feel free to add more variables to see if they are related to time spent, sticking to numeric variables. Were your selected variables related to time spent in the course? 


##### B. APA Formatted Table

While {corrr} is a nice package to quickly create a correlation matrix, you may wish to create one that is ready to be added directly to a dissertation or journal article. {apaTables} is
great for creating more formal forms of output that can be added
directly to an APA-formatted manuscript; it also has functionality for regression and other types of model output. It is not as friendly to {tidyverse} functions; first, we need to select only the variables we wish to correlate.

Then, we can use that subset of the variables as the argument to
the`apa.cor.table()` function.

Run the following code to create a subset of the larger
`data_to_explore` data frame with the variables you wish to correlate, then create a correlation table using `apa.cor.table()`.

```{r messages=FALSE, warning=FALSE}
#install if this is your first time

# read in apatables library
library(apaTables)

data_to_explore_subset <- data_to_explore %>% 
  select(time_spent_hours, proportion_earned, int)

apa.cor.table(data_to_explore_subset)
```
This may look nice, but how to actually add this into a dissertation or article that you might be interested in publishing? 

Read the documentation for `apa.cor.table()` by running `?apa.cor.table()` in the console. Look through the documentation and examples to understand how to output a file with the formatted correlation table, and then run the code to do that with your subset of the `data_to_explore` data frame.

```{r messages=FALSE, warning=FALSE}
apa.cor.table(data_to_explore_subset, filename = "cor-table.doc")
```

You should now see a new Word document in your project folder called `survey-cor-table.doc`. Click on that and you'll be prompted to download from your browser.

##### C. Predict Academic Achievement

##### Linear Regression

In brief, a linear regression model involves
estimating the relationships between one or more *independent variables* with one dependent variable. Mathematically, it can be written like the following.

$$
\operatorname{dependentvar} = \beta_{0} + \beta_{1}(\operatorname{independentvar}) + \epsilon
$$

Does time spent predict grade earned?

The following code estimates a model in which `proportion_earned`, the proportion of points students earned, is the dependent variable. It is predicted by one independent variable, `int`, students' self-reported interest in science.

```{r messages=FALSE, warning=FALSE}
lm(proportion_earned ~ time_spent_hours, 
   data = data_to_explore)
```

The following code estimates a model in which `proportion_earned`, the proportion of points students earned, is the dependent variable. It is predicted by one independent variable, `int`, students' self-reported interest in science.

```{r messages=FALSE, warning=FALSE}
# Add predictor variable for science
lm(proportion_earned ~ time_spent_hours + int, 
   data = data_to_explore)
```


We can see that the intercept is now estimated at 0.44, which tells us that when students' time spent and interest are equal to zero, they are likely fail the course unsurprisingly. Note that that estimate for interest in science is .046, so for every one-unit increase in `int`, we should expect an 5 percentage point increase in their grade.

We can save the output of the function to an object---let's say
`m1`, standing for model 1. We can then use the `summary()` function built into R to view a much more feature-rich summary of the estimated model.

```{r messages=FALSE, warning=FALSE}
# save the model
m1 <- lm(proportion_earned ~ time_spent_hours + int, data = data_to_explore)

```

Run a summary model for the model you just created called, `m1.`
```{r messages=FALSE, warning=FALSE}
#run the summary
summary(m1)

```

##### Summarize predictors
the `summarize()` function from the {dplyr}
package used to create summary statistics such as the mean, standard deviation, or the minimum or maximum of a value. At its core, think of `summarize()` as a function that returns a single value (whether it's a mean, median, standard deviation---whichever!) that summarizes a single column.

```{r messages=FALSE, warning=FALSE}
data_to_explore %>% 
  summarize(mean_interest = mean(int, na.rm = TRUE),
            mean_time = mean(time_spent_hours, na.rm = TRUE))
```

The mean value for interest is quite high. If we multiply the estimate relationship between interest and proportion of points
earned---0.046---by this, the mean interest across all of the
students---we can determine that students' estimate final grade was 0.046 X 4.3, or **0.197**. For hours spent spent, the average students' estimate final grade was 0.0042 X 30.48, or **0.128**.

If we add both 0.197 and 0.128 to the intercept, 0.449, that equals 0.774, or about 77%. In other words, a student with average interest in science who spent an average amount of time in the course earned a pretty average grade.


Let's save this as a nice APA table for possible publication
```{r messages=FALSE, warning=FALSE}
apa.reg.table(m1, filename = "lm-table.doc")
```


```{r messages=FALSE, warning=FALSE}
#install apaTables if this is your first time - do you remember how?

#load packages
library(apaTables)
# use the {apaTables} package to create a nice regression table that could be used for later publication.
apa.reg.table(m1, filename = "lm-table.doc")

```

## Communicate
For your final Your Turn, your goal is to distill our analysis from
above into a simple "data product" designed to illustrate key findings
about changes in the collaboration network over time. For the purposes
of this task, imagine that your audience consists of the developers and
facilitators of the DLT MOOC-Eds who have limited background in SNA and
adapt the following steps accordingly:

1.  **Select.** Select a group  from above, or a new analysis if so motivated, that you think would be interesting or relevant for the target audience and that helps answer our research questions.

2.  **Polish.** Create and polish a data visualization and/or data table to communicate your selected findings.

3.  **Narrate.** Write a brief narrative (2-3 paragraphs) to accompany your visualization and/or table that includes the following:

    -   The question or questions guiding the analysis;

    -   The conclusions you've reached based on our findings;

    -   How your audience might use this information;

    -   How you might revisit or improve upon this analysis in the
        future.
        
#### **👉 Your Turn** **⤵**

Purpose of this case study is to investigate whether time spent in an online course is predictive of student achievement.

To do this we looked at  a simple correlation between time spent and student achievement.

**RQ1**: Do males outperform females in online STEM courses?

```{r warning=FALSE, message=FALSE}
data_viz <- data_to_explore %>%
  select(subject, gender, proportion_earned) %>%  # reduced 
  mutate(subject = recode(subject, 
                          "AnPhA" = "Anatomy",
                          "BioA" = "Biology", 
                          "FrScA" = "Forensics", 
                          "OcnA" =  "Oceanography", 
                          "PhysA" = "Physics")) %>%
  mutate(grade = proportion_earned * 100) %>%
  # filter(!is.na(gender)) %>%
  na.omit() %>% # removed all NAs instead of just those for gender
  group_by(subject, gender) %>% # grouped by subject and gender
  summarise(grade = mean(grade), 
            sd = sd(grade))# calculated mean and sd for grade and saved as grade again  

  ggplot(data_viz, aes(x = subject, y = grade, 
                          fill = gender)) +
  geom_bar(stat = "identity", 
           position = position_dodge()) +
  labs(title = "Do Males out-preform Females in online STEM courses?",
       caption = "Online STEM course performance, why is there still a gender gap?",
       y = "Average Grade",
       x = "Online STEM Course")
```




Want to learn more ways to 

Knit your data product to a desired output format. This may be a simple 
- [HTML](https://bookdown.org/yihui/rmarkdown/html-document.html), 
- [PDF](https://bookdown.org/yihui/rmarkdown/pdf-document.html), or 
- [MS Word](https://bookdown.org/yihui/rmarkdown/word-document.html) file; or something more complex like 
- [HTML5 slides](https://bookdown.org/yihui/rmarkdown/ioslides-presentation.html), 
- [Tufte-style handout](https://bookdown.org/yihui/rmarkdown/tufte-handouts.html),
- [dashboard](https://rmarkdown.rstudio.com/flexdashboard/), or 
- [website](https://bookdown.org/yihui/rmarkdown/rmarkdown-site.html). 

You can publish and share from:

- (e.g., via [RPubs](https://rpubs.com/about/getting-started) or 
- [GitHub](https://towardsdatascience.com/how-to-create-a-free-github-pages-website-53743d7524e1)) 
